Notes on filtration of logics containing {\bf K5}

### 2021-02-13 filtration-stuff/Text.tex
wip

\begin{defin}
  An $n$-normal modal logic is a set of formulas that contains all Boolean tautologies, formulas $\Diamond_i p \lor \Diamond_i q \leftrightarrow \Diamond_i (p \lor q)$ and $\Diamond_i \bot \leftrightarrow \bot$ for $i \leq n$, and is closed under modus ponens, substitution, and monotonicity: from $\varphi \rightarrow \psi$ infer $\Diamond_i \varphi \rightarrow \Diamond_i \psi$ for $i \leq n$.
\end{defin}

\begin{defin} An $n$-Kripke model is a triple $\mathcal{M} = \langle W, R_1, \dots, R_n, \vartheta \rangle$, where $R_i \subseteq W \times W$, $\vartheta : \operatorname{PV} \to 2^W$, and the connectives have the following semantics:

  \begin{enumerate}
    \item $\mathcal{M}, w \models p \Leftrightarrow w \in \vartheta(p)$
    \item $\mathcal{M}, w \models \not \varphi \Leftrightarrow \mathcal{M}, w \nvDash \varphi$
    \item $\mathcal{M}, w \models \varphi \lor \psi \Leftrightarrow \mathcal{M}, w \models \varphi \text{ or } \mathcal{M}, w \models \psi$
    \item $\mathcal{M}, w \models \Diamond_i \varphi \Leftrightarrow \exists v \in R_i(w) \: \mathcal{M}, v \models \varphi$
  \end{enumerate}
\end{defin}

@@

\section{Finite ``canonical`` models}

@@

\begin{defin} Let $\varphi$ be a modal formula, $\mathcal{L}$ a logic that admits strict filtration, and $\mathcal{M}$ an $\mathcal{L}$-model that refutes $\varphi$. A \emph{filtration diagram} of a model $\widehat{\mathcal{M}}$ (a filtration of $\mathcal{M}$ through $\varphi$) is a formula $\operatorname{FDiagram}(\mathcal{M}_{\mathcal{L}}^{\operatorname{Sub}(\varphi)}) = \bigwedge \Gamma \wedge \bigwedge \Delta$, where
  \begin{itemize}
    \item $\Gamma =
    \{ \psi \in \operatorname{Sub}(\varphi) \: | \: \exists w \in \underline{\mathcal{M}} \: \widehat{M}, \hat{w} \models \psi \}$
    \item $\Delta = \{ \neg \psi \in \operatorname{Sub}(\varphi) \: | \: \psi \in \operatorname{Sub}(\varphi) \: \& \: \mathcal{M} \not\models \psi \}$
  \end{itemize}
\end{defin}

@@

\begin{theorem} Let $\mathcal{L}$ be a complete normal modal logic, then the following are equivalent:
  \begin{enumerate}
    \item $\mathcal{L}$ admits strict filtrations
    \item If $\varphi \not\in \mathcal{L}$, there exists a finite ``canonical'' model $\mathcal{M}_{\mathcal{L}}^{\varphi}$ such that
    $\mathcal{M}_{\mathcal{L}}^{\varphi} \models \mathcal{L}$ and
    $\mathcal{M}_{\mathcal{L}}^{\varphi}$ refutes $\varphi$.
  \end{enumerate}
\end{theorem}

@@

    Let $\varphi \not\in \mathcal{L}$, there exists a model $\mathcal{M}$ that refutes $\varphi$, so there exists $w \in \underline{\mathcal{M}}$ such that $\mathcal{M}, w \not\models \varphi$. But $\mathcal{L}$ admits strict filtration, so there exists a model $\widehat{\mathcal{M}} = \langle \widehat{W}, \widehat{R}, \widehat{\vartheta} \rangle$, where $\widehat{W} = W / \sim_{\varphi}$ and $\hat{R}^{min} \subseteq \widehat{R} \subseteq \hat{R}^{max}$. In particular, $\widehat{\mathcal{M}}, \widehat{w} \not\models \varphi$.

@@

    Let $\mathcal{M}_{\mathcal{L}}^{\varphi}$ be a finite ``canonical'' model that refutes $\varphi$.

    Consider $W_{\mathcal{L}}^{\varphi} / \sim_{\operatorname{Sub}(\varphi)}$. Clearly $W_{\mathcal{L}}^{\varphi} = W_{\mathcal{L}}^{\varphi} / \sim_{\operatorname{Sub}(\varphi)}$ since

    \begin{center}
      $\Delta_1 \sim_{\operatorname{Sub}(\varphi)} \Delta_2$ iff $\forall \psi \in \operatorname{Sub}(\varphi) \:\: W_{\mathcal{L}}^{\varphi}, \Delta_1 \models \psi \Leftrightarrow W_{\mathcal{L}}^{\varphi}, \Delta_2 \models \psi$ iff $\Delta_1 = \Delta_2$.
    \end{center}

    $R_{\mathcal{L}}^{\varphi}$ also satisfies the required definition.
    Suppose $\Box \psi \in \varphi$ and
    $\mathcal{M}_{\mathcal{L}}^{\varphi}, \Delta_1 \models \Box \psi$ and
    $\Delta_1 R_{\mathcal{L}}^{\varphi} \Delta_2$. Clearly that $\psi \in \Delta_2$ by the definition
    of a finite ``canonical'' model.

    Therefore, $\mathcal{L}$ admits strict filtrations since $\mathcal{M}_{\mathcal{L}}^{\varphi}$ is already finite. In particular, ${\bf K}5$ fails to have the finite ``canonical'' model property that follows from the contraposition of this statement and Theorem~\ref{sad}.

  \end{enumerate}
\end{proof}

>>>>

\begin{defin}
  A normal modal logic is a set of formulas that contains all Boolean tautologies,
  formulas $\Diamond p \lor \Diamond q \leftrightarrow \Diamond (p \lor q)$ and
  $\Diamond \bot \leftrightarrow \bot$, and is closed under Modus Ponens, Substitution, and
  Monotonicity:
  from $\varphi \rightarrow \psi$ infer $\Diamond \varphi \rightarrow \Diamond \psi$.
\end{defin}

\begin{defin} An Kripke model is a triple $\mathcal{M} = \langle W, R, \vartheta \rangle$,
  where $R \subseteq W \times W$, $\vartheta : \operatorname{PV} \to 2^W$, and
  the connectives have the following semantics:

  \begin{enumerate}
    \item $\mathcal{M}, w \models p \Leftrightarrow w \in \vartheta(p)$
    \item $\mathcal{M}, w \models \neg \varphi \Leftrightarrow \mathcal{M}, w \nvDash \varphi$
    \item $\mathcal{M}, w \models \varphi \lor \psi \Leftrightarrow \mathcal{M}, w \models \varphi \text{ or } \mathcal{M}, w \models \psi$
    \item $\mathcal{M}, w \models \Diamond \varphi \Leftrightarrow \exists v \in R(w) \: \mathcal{M}, v \models \varphi$
  \end{enumerate}
\end{defin}

\begin{defin}
  Let $\mathcal{M} = \langle W, R, \vartheta \rangle$ be a Kripke model.
  A Kripke model $\mathcal{M}' = \langle W', R', \vartheta' \rangle$ is a generated submodel of
  $\mathcal{M}$, where:
  \begin{enumerate}
    \item $\emptyset \neq W' \subseteq W$ and $W'$ is $R$-closed, i.e., for each $u \in W'$ and $v \in W$,
    $u R v$ implies $v \in W'$
    \item $R' = R \cap W' \times W'$
    \item for every propositional variable $p$, $\vartheta(p)' = \vartheta(p) \cap W'$
  \end{enumerate}
\end{defin}

\begin{fact}
  Let $\mathcal{M} = \langle W, R, \vartheta \rangle$ be a Kripke model and
  $\mathcal{M}' = \langle W', R', \vartheta' \rangle$ its generated submodel,
  then for each $w \in W'$
  \begin{center}
    $\mathcal{M}, w \models \varphi$ iff $\mathcal{M}', w \models \varphi$
  \end{center}
\end{fact}

@@

\section{Finite ``canonical'' models}

@@

\begin{defin} Let $\varphi$ be a modal formula, $\mathcal{L}$ a logic that admits strict filtration, and $\mathcal{M}$ an $\mathcal{L}$-model that refutes $\varphi$. A \emph{filtration diagram} of a model $\widehat{\mathcal{M}}$ (a filtration of $\mathcal{M}$ through $\varphi$) is a formula $\operatorname{FDiagram}(\mathcal{M}_{\mathcal{L}}^{\operatorname{Sub}(\varphi)}) = \bigwedge \Gamma \wedge \bigwedge \Delta$, where
  \begin{itemize}
    \item $\Gamma =
    \{ \psi \in \operatorname{Sub}(\varphi) \: | \: \exists w \in \underline{\mathcal{M}} \: \widehat{M}, \hat{w} \models \psi \}$
    \item $\Delta = \{ \neg \psi \in \operatorname{Sub}(\varphi) \: | \: \psi \in \operatorname{Sub}(\varphi) \: \& \: \mathcal{M} \nvDash \psi \}$
  \end{itemize}
\end{defin}

@@

\begin{theorem} Let $\mathcal{L}$ be a canonical logic, then the following are equivalent:
  \begin{enumerate}
    \item $\mathcal{L}$ admits strict filtrations
    \item If $\varphi \not\in \mathcal{L}$, there exists a finite ``canonical'' model $\mathcal{M}_{\mathcal{L}}^{\varphi}$ such that
    $\mathcal{M}_{\mathcal{L}}^{\varphi} \models \mathcal{L}$ and
    $\mathcal{M}_{\mathcal{L}}^{\varphi}$ refutes $\varphi$.
  \end{enumerate}
\end{theorem}

@@

    Let $\varphi \not\in \mathcal{L}$, there exists a model $\mathcal{M}$ that refutes $\varphi$, so there exists $w \in \underline{\mathcal{M}}$ such that $\mathcal{M}, w \nvDash \varphi$. But $\mathcal{L}$ admits strict filtration, so there exists a model $\widehat{\mathcal{M}} = \langle \widehat{W}, \widehat{R}, \widehat{\vartheta} \rangle$, where $\widehat{W} = W / \sim_{\varphi}$ and $\hat{R}^{min} \subseteq \widehat{R} \subseteq \hat{R}^{max}$. In particular, $\widehat{\mathcal{M}}, \widehat{w} \nvDash \varphi$.

@@

    Let $\varphi \not\in \mathcal{L}$ and $\mathcal{M}_{\mathcal{L}}^{\varphi} = \langle W_{\mathcal{L}}^{\varphi}, R_{\mathcal{L}}^{\varphi}, \vartheta_{\mathcal{L}}^{\varphi} \rangle$
    be a finite ``canonical'' model that refutes $\varphi$.
    Let $\mathcal{M}$ be an $\mathcal{L}$-model that also refutes $\varphi$.
    Let us show that there exists a model $\widehat{\mathcal{M}}$ such that
    $\widehat{\mathcal{M}}$ is a
    strict filtration of $\mathcal{M}$ through $\operatorname{Sub}(\varphi)$.

    Let $\sim_{\operatorname{Sub}(\varphi)}$ be an equivalence relation on $\underline{\mathcal{M}}$ defined as usual:

    \begin{center}
      $u \sim_{\operatorname{Sub}(\varphi)} v$ iff $\forall \psi \in \operatorname{Sub}(\varphi)$
      $\mathcal{M}, u \models \psi \Leftrightarrow \mathcal{M}, v \models \psi$
    \end{center}

    Let $[u] \in W/\sim_{\operatorname{Sub}(\varphi)}$, where $W = \underline{\mathcal{M}}$,
    consider its truth set $||[u]|| = \{ \psi \in \operatorname{Sub}(\varphi) \: | \: \mathcal{M}, u \models \psi \}$.
    $||[u]||$ is clearly a finite consistent theory and it consists in some $\Delta \in W_{\mathcal{L}}^{\varphi}$ by the finite analogue of Lindenbaum's lemma. Such a $\Delta$ is clearly unique as well.
    Consider a map
    \begin{center}
    $\iota : W/\sim_{\operatorname{Sub}(\varphi)} \to W_{\mathcal{L}}^{\varphi}$ with
    $\iota : [u] \mapsto \Delta \supseteq ||[u]||$
    \end{center}

    Consider the set $\iota(W/\sim_{\operatorname{Sub}(\varphi)})$ denoted as $\operatorname{Im}(\iota)$. Let us show that $\operatorname{Im}(\iota)$ is $R_{\mathcal{L}}^{\varphi}$-closed.

    Let $\iota([u]) R_{\mathcal{L}}^{\varphi} \Delta$, let us show that $\Delta \in \iota(W/\sim_{\operatorname{Sub}(\varphi)})$, that is, $\Delta = \iota([v])$ for some $v \in W$.

    Consider $\Delta$ as a finite ultrafilter. Thus, $\Delta$ is principal and it 's generated by $\psi_1 \land \dots \land \psi_n$, where each $\psi_i \in \operatorname{Sub}(\varphi)$. So $\Diamond (\psi_1 \land \dots \land \psi_n) \in \iota([u])$.
    Then $\mathcal{M}, u \models \Diamond (\psi_1 \land \dots \land \psi_n)$. That implies $\mathcal{M}, v \models \psi_1 \land \dots \land \psi_n$ for some $v \in R(u)$.
    This is clearly a desired $v$ since $||[v]|| \subseteq \Delta$ and such a $\Delta$ is unique.

    Thus, $\langle \operatorname{Im}(\iota), R_{\mathcal{L}}^{\varphi} \cap \operatorname{Im}(\iota) \times \operatorname{Im}(\iota), {\vartheta'}_{\mathcal{L}}^{\varphi} \rangle$, where ${\vartheta'}_{\mathcal{L}}^{\varphi}(p) = \vartheta_{\mathcal{L}}^{\varphi}(p) \cap \operatorname{Im}(\iota)$ is a generated submodel of $\mathcal{M}_{\mathcal{L}}^{\varphi}$.

    Consider a model $\widehat{\mathcal{M}} = \langle W/\sim_{\operatorname{Sub}(\varphi)}, \widehat{R}, \widehat{\vartheta} \rangle$, where $\widehat{R}$ is defined as
    \begin{center}
      $[u] \widehat{R} [v]$ iff $||[u]|| {R'}_{\mathcal{L}}^{\varphi} ||[v]||$, where ${R'}_{\mathcal{L}}^{\varphi} R_{\mathcal{L}}^{\varphi} \cap \operatorname{Im}(\iota) \times \operatorname{Im}(\iota)$.
    \end{center}
    Let us prove that $\widehat{\mathcal{M}}$ is a filtration of $\mathcal{M}$ through $\operatorname{Sub}(\varphi)$.

    Let us show that $R_{min} \subseteq \widehat{R} \subseteq R_{max}$.

    Suppose that there are $u' \in [u]$ and $v' \in [v]$ with $u' R v'$. Let us show that
    $[u] \widehat{R} [v]$. That is, let us show that $\Diamond \iota(v) \subseteq \iota(v)$.
    But if $\mathcal{M}, v' \models \psi$ for $\Diamond \psi \in \operatorname{Sub}(\varphi)$, then
    $\mathcal{M}, u' \models \Diamond \psi$. Recall that $u' \sim_{\operatorname{Sub}(\varphi)} u$ and $v' \sim_{\operatorname{Sub}(\varphi)} v$, so we have $\Diamond ||[v]|| \subseteq ||[u]||$.
    The same inclusion clearly holds for the maximal extensions of $\Diamond ||[v]||$ and $||[u]||$.

    The second inclusion follows from the definition of a relation in a finite ``canonical'' model.
  \end{enumerate}
\end{proof}

In particular, ${\bf K}5$ fails to have the finite ``canonical'' model property that
follows from the contraposition of the theorem above and Theorem~\ref{sad}.
